---
title: "MSEA"
author: "Margaret Janiczek"
date: "2025-01-29"
output:
  html_document:
    df_print: paged
    toc: TRUE
    toc_float: TRUE
    number_sections: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
#BiocManager::install("fgsea")
library(fgsea)
library(ggplot2)
library(tidyr)
library(dplyr)
library(purrr)
```



```{r}
#now load metabolite group info

#
load("processed_met_key.Rdata")

library(stringr)

met_key <- ukb_key_metab %>%
  filter(!str_detect(col.name, "qc_flag"))%>%
  filter(field.showcase %in% 23400:24000) %>%
  filter(col.type == "Continuous") %>%
  filter(str_detect(field.tab, ".0.0")) %>%
  mutate(pathway = ifelse(
    Group == "Lipoprotein subclasses", Subgroup,Group
  ))

#met_key$pathway[is.na(met_key$pathway)] <- "Other"

# make list of mets per met group

met_key_select <- met_key %>%
  filter(!(pathway %in% c("Ratios","Lipoprotein particle sizes", 
           "Apolipoproteins", "Fluid balance", "Inflammation"))) %>%
  mutate(pathway = as.factor(str_split_i(pathway, '\\(', 1))) %>%
  mutate(pathway = fct_relevel(pathway, "Chylomicrons and extremely large VLDL ", "Very large VLDL ", "Large VLDL ", "Medium VLDL ", "Small VLDL ", "Very small VLDL ", "Amino acids", "Ketone bodies", "Triglycerides", "Small LDL", "Medium LDL", "Large LDL", "Total lipids", "Phospholipids", "Glycolysis related metabolites", "Cholesteryl esters", "Lipoprotein particle concentrations", "Cholesterol", "Fatty acids", "Free cholesterol", "IDL ", "Other lipids", "Small HDL ", "Medium HDL ", "Large HDL ", "Very large HDL "))

met_list <- split(met_key_select$col.name,met_key_select$pathway)

# res81 <- read.csv("validated_unique_mets_81_details.csv")
# 
# res812 <- read.csv("results_mod1_bothphases.csv") %>%
#   #filter(metabolite %in% res81$metabolite) %>%
#   arrange(estimate)

#load("results_mod2_bothphases.Rdata")

load("m1_msea_mod.Rdata")
load("m2_msea_mod.Rdata")
load("m3_msea_mod.Rdata")
```

# MSEA Age Category <50

## M1

Using estimates from M1 age <50 (both phases)

```{r}


ranks_lt50_m1 <- all_strat_M1_msea %>% 
   filter(age_cat == "<50")  %>%
   pull(estimate)

names(ranks_lt50_m1) <- all_strat_M1_msea %>% filter(age_cat == "<50") %>% pull(metabolite)

fgsea_res_m1_lt50 <- fgsea(
  pathways = test,
  stats = ranks_lt50_m1
) %>%
  mutate(age_cat = "<50",
         model = "M1")

head(fgsea_res_m1_lt50)

topPathwaysUp <- fgsea_res_m1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_m1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50_m1, fgsea_res_m1_lt50, gseaParam = 0.5)
```




## M2

Using estimates from M2 age <50

```{r}


ranks_lt50_m2 <- all_strat_M2_msea %>% 
   filter(age_cat == "<50")  %>%
   pull(estimate)

names(ranks_lt50_m2) <- all_strat_M2_msea %>% filter(age_cat == "<50") %>% pull(metabolite)

fgsea_res_m2_lt50 <- fgsea(
  pathways = test,
  stats = ranks_lt50_m2
) %>%
  mutate(age_cat = "<50",
         model = "M2")

head(fgsea_res_m2_lt50)

#plotEnrichment(test[["Large HDL (average diameter 12.1 nm)"]],ranks_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```



# MSEA Age Category 50-59

## M1

Using estimates from M1 age 50-59

```{r}


ranks_lt59_m1 <- all_strat_M1_msea %>% 
   filter(age_cat == "50-59")  %>%
   pull(estimate)

names(ranks_lt59_m1) <- all_strat_M1_msea %>% filter(age_cat == "50-59") %>% pull(metabolite)

fgsea_res_m1_lt59 <- fgsea(
  pathways = test,
  stats = ranks_lt59_m1
) %>%
  mutate(age_cat = "50-59",
         model = "M1")

head(fgsea_res_m1_lt50)

plotEnrichment(test[["Large HDL (average diameter 12.1 nm)"]],ranks_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```

## M2

Using estimates from M2 age 50-59

```{r}


ranks_lt59_m2 <- all_strat_M2_msea %>% 
   filter(age_cat == "50-59")  %>%
   pull(estimate)

names(ranks_lt59_m2) <- all_strat_M2_msea %>% filter(age_cat == "50-59") %>% pull(metabolite)

fgsea_res_m2_lt59 <- fgsea(
  pathways = test,
  stats = ranks_lt59_m2
) %>%
  mutate(age_cat = "50-59",
         model = "M2")

head(fgsea_res_p1_lt50)

plotEnrichment(test[["Large HDL (average diameter 12.1 nm)"]],ranks_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```




# MSEA Age Category 60-69

## M1

Using estimates from M1 age 60-69

```{r}


ranks_lt69_m1 <- all_strat_M1_msea %>% 
   filter(age_cat == "60-69")  %>%
   pull(estimate)

names(ranks_lt69_m1) <- all_strat_M1_msea %>% filter(age_cat == "60-69") %>% pull(metabolite)

fgsea_res_m1_lt69 <- fgsea(
  pathways = test,
  stats = ranks_lt69_m1
) %>%
  mutate(age_cat = "60-69",
         model = "M1")

head(fgsea_res_p1_lt50)

plotEnrichment(test[["Large HDL (average diameter 12.1 nm)"]],ranks_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```

## M2

Using estimates from M2 age 60-69

```{r}


ranks_lt69_m2 <- all_strat_M2_msea %>% 
   filter(age_cat == "60-69")  %>%
   pull(estimate)

names(ranks_lt69_m2) <- all_strat_M2_msea %>% filter(age_cat == "60-69") %>% pull(metabolite)

fgsea_res_m2_lt69 <- fgsea(
  pathways = test,
  stats = ranks_lt69_m2
) %>%
  mutate(age_cat = "60-69",
         model = "M2")

head(fgsea_res_p1_lt50)

plotEnrichment(test[["Large HDL (average diameter 12.1 nm)"]],ranks_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```


# MSEA Age Category 70-85

## M1 

Using estimates from M1 age 70-85

```{r}


ranks_lt85_m1 <- all_strat_M1_msea %>% 
   filter(age_cat == "70-85")  %>%
   pull(estimate)

names(ranks_lt85_m1) <- all_strat_M1_msea %>% filter(age_cat == "70-85") %>% pull(metabolite)

fgsea_res_m1_lt85 <- fgsea(
  pathways = test,
  stats = ranks_lt85_m1
) %>%
  mutate(age_cat = "70-85",
         model = "M1")

head(fgsea_res_p1_lt50)

plotEnrichment(test[["Large HDL (average diameter 12.1 nm)"]],ranks_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```

## M2

Using estimates from M2 age 70-85

```{r}


ranks_lt85_m2 <- all_strat_M2_msea %>% 
   filter(age_cat == "70-85")  %>%
   pull(estimate)

names(ranks_lt85_m2) <- all_strat_M2_msea %>% filter(age_cat == "70-85") %>% pull(metabolite)

fgsea_res_m2_lt85 <- fgsea(
  pathways = test,
  stats = ranks_lt85_m2
) %>%
  mutate(age_cat = "70-85",
         model = "M2")

head(fgsea_res_p1_lt50)

plotEnrichment(test[["Large HDL (average diameter 12.1 nm)"]],ranks_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```

## M3

Using estimates from M3

```{r}
all_strat_select <- all_strat_both_phases_long_m3 %>%
  filter(metabolite %in% met_key_select$col.name)

ranks_lt50_m3 <- all_strat_both_phases_long_m3 %>% 
  filter(phase == "Phase I") %>%
   filter(age_cat == "<50")  %>%
   pull(estimate)

names(ranks_lt50_m3) <- all_strat_select %>% 
  filter(phase == "Phase I") %>% filter(age_cat == "<50") %>% pull(metabolite)

fgsea_res_m3_lt50 <- fgsea(
  pathways = met_list,
  stats = ranks_lt50_m3
) %>%
  mutate(age_cat = "<50",
         model = "M3")

ranks_lt59_m3 <- all_strat_select %>% 
  filter(phase == "Phase I") %>%
   filter(age_cat == "50-59")  %>%
   pull(estimate)

names(ranks_lt59_m3) <- all_strat_select %>% 
  filter(phase == "Phase I") %>% filter(age_cat == "50-59") %>% pull(metabolite)

fgsea_res_m3_lt59 <- fgsea(
  pathways = met_list,
  stats = ranks_lt59_m3
) %>%
  mutate(age_cat = "50-59",
         model = "M3")

ranks_lt69_m3 <- all_strat_select %>% 
  filter(phase == "Phase I") %>% 
   filter(age_cat == "60+")  %>%
   pull(estimate)

names(ranks_lt69_m3) <- all_strat_select %>% 
  filter(phase == "Phase I") %>% filter(age_cat == "60+") %>% pull(metabolite)

fgsea_res_m3_lt69 <- fgsea(
  pathways = met_list,
  stats = ranks_lt69_m3
) %>%
  mutate(age_cat = "60+",
         model = "M3")

topPathwaysUp <- fgsea_res_m3_lt59 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_m3_lt59 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))
```


```{r}
all_msea_res <- bind_rows(list(fgsea_res_m3_lt50, fgsea_res_m3_lt59, fgsea_res_m3_lt69))
```

```{r}
# forest for msea 

# trying loop to get plot for all the subgroups

met_cats <- unique(all_msea_res$pathway)

models <- unique(all_msea_res$model)

map_res <- data.frame(
  models = models) %>%
  mutate(data = map(models, ~ all_msea_res %>% 
                      filter(model == .x)%>%
                      arrange(NES) %>%
                      complete(age_cat,pathway,
                               fill = list(NES = 0))))
library(forcats)

map2(.x = map_res$data, .y = models, 
                        ~ggplot(.x,
       aes(x = pathway, y = NES, col = age_cat, fill = age_cat)) +
        geom_col(position = "dodge") +
        coord_flip() +
        theme_minimal() +
        theme() +
        ggtitle(.y))

res_fig <- all_msea_res %>%
 # arrange(NES) %>%
  #filter(padj <0.05)%>%
                      complete(age_cat,pathway,model,
                               fill = list(NES = 0)) %>%
  mutate(pathways = fct_relevel(pathway, "Chylomicrons and extremely large VLDL ", "Very large VLDL ", "Large VLDL ", "Medium VLDL ", "Small VLDL ", "Very small VLDL ", "Amino acids", "Ketone bodies", "Triglycerides", "Small LDL ", "Medium LDL ", "Large LDL ", "Total lipids", "Phospholipids", "Glycolysis related metabolites", "Cholesteryl esters", "Lipoprotein particle concentrations", "Cholesterol", "Fatty acids", "Free cholesterol", "IDL ", "Other lipids", "Small HDL ", "Medium HDL ", "Large HDL ", "Very large HDL "))%>%
  mutate(significant = ifelse(padj <0.05, 1, 0))

ggplot(res_fig,
       aes(x = pathways, y = NES, col = age_cat, fill = age_cat)) +
        geom_col(position = "dodge") +
        geom_text(aes(label = ifelse(significant, "*", ""), group = age_cat), color = "black", 
            position = position_dodge(width = .9), size = 20 / .pt)+
        coord_flip() +
        theme_minimal() +
        theme() +
        facet_grid(~model)
```

