---
title: "first occurrences process"
output: pdf_document
date: "2023-08-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
load("/rawdata/UKBB/first_occurences/first_occ_processed.Rdata")

# grab columns of occurrence
event_key <- first_occ_key %>% filter(str_detect(col.name, "source"))
eventdate_key <- first_occ_key %>% filter(str_detect(col.name, "date"))

event <- first_occ %>%
  select(eid, all_of(event_key$col.name))

event_date <- first_occ %>% select(eid, all_of(eventdate_key$col.name))

# go to longer
event_long <- event %>%
  pivot_longer(!eid, values_drop_na = TRUE) %>%
  group_by(eid) %>%
  mutate(id = row_number())

# grab date of first occurence
eventdate_long <- event_date %>%
  pivot_longer(!eid, values_drop_na = TRUE) %>%
  group_by(eid) %>%
  mutate(id = row_number())

#load date of attending assessment center first
load("/rawdata/UKBB/Metabolomics_Apr2023/icd10.Rdata")
dat_first_event <- first_date %>%
  right_join(eventdate_long) %>%
  mutate(afterassessment_fo = value > date_of_attending_assessment_centre_f53_0_0,
         beforeassessment_fo = value < date_of_attending_assessment_centre_f53_0_0) ## this should give prevalence/occurence before baseline

# filter icd10 codes for various conditions

event_long <- event_long %>%
  mutate(
    cad = ifelse(str_detect(name, c("i20|i21|i22|i23|i24|i25|z951|z955")),1,0),
    afib = ifelse(str_detect(name, c("i48")), 1, 0),
    stroke = ifelse(str_detect(name, c("i60|i61|i629|i63|i64|i678|i690|i693|g951|h341|h342|s066")),1,0),
    hypertension = ifelse(
      str_detect(
        name, c("i10|i11|i12|i13|i15|o10")
      ), 1,0
    ),
    diabetes_other = ifelse(str_detect(
      name, c("e10|e11|e13|e14") # separated out other diabetes 11/22/23
    ), 1, 0),
    diabetes_t2 = ifelse(str_detect(
      name, c("e11")
    ), 1, 0)
  )

test <- event_long %>%
  pivot_longer(!(eid:id), names_to = "disease_fo", 
values_to = "has_disease_fo",
values_drop_na = FALSE) %>%
  group_by(eid, disease_fo, has_disease_fo) %>%
  slice(1) # just grab first instance of disease

test3 <- test %>%
  select(eid, id, source_of_report = value, disease_fo, has_disease_fo) %>%
  filter(has_disease_fo == 1) %>%
  group_by(eid, disease_fo) %>%
  slice(1)

final_fo <- dat_first_event %>%
  right_join(test3) #%>%
  #filter(afterassessment_fo == TRUE) # don't filter this and then you can have both variables before and after assessment

final_fo <- final_fo %>%
  mutate(disease_source = paste0(disease_fo, source_of_report))

# saving a new version as of Jan 2024 that has variable for before assessment
save(final_fo, file = "/rawdata/UKBB/Metabolomics_jul2023/first_occ_all_long_010524.Rdata")

library(ggplot2)
final_fo %>%
  filter(has_disease_fo == 1) %>%
  ggplot(aes(disease_fo)) +
  geom_bar(aes(fill = source_of_report)) +
  theme_bw() +
  ylab("Percent") +
  xlab("Disease")

length(unique(final_fo$eid))

# get set of unique patients that have one of the diseases after baseline

final2_fo <- final_fo %>%
  select(eid, disease_fo, has_disease_fo) %>%
  #mutate(source_of_report = as.character(source_of_report)) %>%
  pivot_wider(names_from = disease_fo, values_from = c(has_disease_fo), values_fill = 0) %>%
  ungroup() %>%
  group_by(eid) %>%
  mutate(id = row_number())


#load metab pheno processed data

load("/rawdata/UKBB/Metabolomics_jul2023/dat_subset_metab.Rdata")

dat_sub_metab2 <- dat_sub_metab2 %>%
  left_join(final2_fo)

dat_sub_metab2$cad_fo[is.na(dat_sub_metab2$cad_fo)] <- 0
dat_sub_metab2$afib_fo[is.na(dat_sub_metab2$afib_fo)] <- 0
dat_sub_metab2$stroke_fo[is.na(dat_sub_metab2$stroke_fo)] <- 0
dat_sub_metab2$hypertension_fo[is.na(dat_sub_metab2$hypertension_fo)] <- 0
dat_sub_metab2$diabetes_t2_fo[is.na(dat_sub_metab2$diabetes_t2_fo)] <- 0
dat_sub_metab2$cad[is.na(dat_sub_metab2$cad)] <- 0
dat_sub_metab2$afib[is.na(dat_sub_metab2$afib)] <- 0
dat_sub_metab2$stroke[is.na(dat_sub_metab2$stroke)] <- 0
dat_sub_metab2$hypertension[is.na(dat_sub_metab2$hypertension)] <- 0
dat_sub_metab2$diabetes_t2[is.na(dat_sub_metab2$diabetes_t2)] <- 0
dat_sub_metab2$diabetes_other[is.na(dat_sub_metab2$diabetes_other)] <- 0

var_label(dat_sub_metab2) <- outcomedatlist_fo
var_label(dat_sub_metab2) <- outcomedatlist
var_label(dat_sub_metab2) <- covardatlist

save(dat_sub_metab2, file = "/rawdata/UKBB/Metabolomics_jul2023/dat_subset_metab_112223.Rdata")


```

```{r}
tab_disease <- dat_sub_metab2 %>%
  select(sex_f31_0_0, metabolite_status, all_of(names(unlist(outcomedatlist))), all_of(names(unlist(outcomedatlist_fo))))

metab_tab_phase1_sex <- tab_disease %>%
  filter(metabolite_status == "Phase 1") %>%
  select(-metabolite_status) %>%
  tbl_summary(by = sex_f31_0_0, missing = "no") %>%
  bold_labels() %>%
  add_n() %>%
  add_overall()

all_tab_phase1_outcome <- tab_disease %>%
  select(-metabolite_status) %>%
  tbl_summary(by = sex_f31_0_0, missing = "no") %>%
  bold_labels() %>%
  add_n() %>%
  add_overall()


test <- first_occ_key %>%
  filter(str_detect(col.name, "i61"))
```


** First occurrences data only had codes for I10 (primary hypertension) and I15 (secondary hypertension). Our original code included I11 (hypertensive heart disease with heart failure), I12 (Hypertensive chronic kidney disease), I13 (hypertensive heart and chronic kidney disease) and O10 (pre-existing hypertension complicating pregnancy). This accounts for the discrepancy between first occurrences and our original ICD10 data.

** First occurrences data for stroke included I60, I61, I63, I64. But it only included broad categories like I62 - not specific conditions like I62.9, I67.8, I69.0.  

** First occurrences for "coronary artery disease" only included I21-25 (not Z951 or Z955)

** Diabetes and Atrial Fibrillation had same codes present in both places. 

