---
title: "MSEA"
author: "Margaret Janiczek"
date: "2024-01-03"
output:
  html_document:
    df_print: paged
    toc: TRUE
    toc_float: TRUE
    number_sections: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
#BiocManager::install("fgsea")
library(fgsea)
library(ggplot2)
library(tidyr)
library(dplyr)
library(purrr)
```

```{r}
# example
data(examplePathways)
data(exampleRanks)
set.seed(42)

fgseaRes <- fgsea(pathways = examplePathways,
                  stats = exampleRanks,
                  minSize = 15,
                  maxSize = 500)
```

```{r}
#now load metabolite group info

#
load("processed_met_key.Rdata")

library(stringr)

met_key <- ukb_key_metab %>%
  filter(!str_detect(col.name, "qc_flag"))%>%
  filter(field.showcase %in% 23400:24000) %>%
  filter(col.type == "Continuous") %>%
  filter(str_detect(field.tab, ".0.0")) %>%
  mutate(pathway = ifelse(
    Group == "Lipoprotein subclasses", Subgroup,Group
  ))

met_key$pathway[is.na(met_key$pathway)] <- "Other"

# make list of mets per met group
test <- split(met_key$col.name,met_key$pathway)

# res81 <- read.csv("validated_unique_mets_81_details.csv")
# 
# res812 <- read.csv("results_mod1_bothphases.csv") %>%
#   #filter(metabolite %in% res81$metabolite) %>%
#   arrange(estimate)

#load("results_mod2_bothphases.Rdata")

load("m1_msea_mod.Rdata")
load("m2_msea_mod.Rdata")
```

# MSEA Age Category <50

## M1

Using estimates from M1 age <50 (both phases)

```{r}


ranks_lt50_m1 <- all_strat_M1_msea %>% 
   filter(age_cat == "<50")  %>%
   pull(estimate)

names(ranks_lt50_m1) <- all_strat_M1_msea %>% filter(age_cat == "<50") %>% pull(metabolite)

fgsea_res_m1_lt50 <- fgsea(
  pathways = test,
  stats = ranks_lt50_m1
) %>%
  mutate(age_cat = "<50",
         model = "M1")

head(fgsea_res_p1_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```




## M2

Using estimates from M2 age <50

```{r}


ranks_lt50_m2 <- all_strat_M2_msea %>% 
   filter(age_cat == "<50")  %>%
   pull(estimate)

names(ranks_lt50_m2) <- all_strat_M2_msea %>% filter(age_cat == "<50") %>% pull(metabolite)

fgsea_res_m2_lt50 <- fgsea(
  pathways = test,
  stats = ranks_lt50_m2
) %>%
  mutate(age_cat = "<50",
         model = "M2")

head(fgsea_res_p1_lt50)

plotEnrichment(test[["Large HDL (average diameter 12.1 nm)"]],ranks_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```



# MSEA Age Category 50-59

## M1

Using estimates from M1 age 50-59

```{r}


ranks_lt59_m1 <- all_strat_M1_msea %>% 
   filter(age_cat == "50-59")  %>%
   pull(estimate)

names(ranks_lt59_m1) <- all_strat_M1_msea %>% filter(age_cat == "50-59") %>% pull(metabolite)

fgsea_res_m1_lt59 <- fgsea(
  pathways = test,
  stats = ranks_lt59_m1
) %>%
  mutate(age_cat = "50-59",
         model = "M1")

head(fgsea_res_p1_lt50)

plotEnrichment(test[["Large HDL (average diameter 12.1 nm)"]],ranks_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```

## M2

Using estimates from M2 age 50-59

```{r}


ranks_lt59_m2 <- all_strat_M2_msea %>% 
   filter(age_cat == "50-59")  %>%
   pull(estimate)

names(ranks_lt59_m2) <- all_strat_M2_msea %>% filter(age_cat == "50-59") %>% pull(metabolite)

fgsea_res_m2_lt59 <- fgsea(
  pathways = test,
  stats = ranks_lt59_m2
) %>%
  mutate(age_cat = "50-59",
         model = "M2")

head(fgsea_res_p1_lt50)

plotEnrichment(test[["Large HDL (average diameter 12.1 nm)"]],ranks_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```




# MSEA Age Category 60-69

## M1

Using estimates from M1 age 60-69

```{r}


ranks_lt69_m1 <- all_strat_M1_msea %>% 
   filter(age_cat == "60-69")  %>%
   pull(estimate)

names(ranks_lt69_m1) <- all_strat_M1_msea %>% filter(age_cat == "60-69") %>% pull(metabolite)

fgsea_res_m1_lt69 <- fgsea(
  pathways = test,
  stats = ranks_lt69_m1
) %>%
  mutate(age_cat = "60-69",
         model = "M1")

head(fgsea_res_p1_lt50)

plotEnrichment(test[["Large HDL (average diameter 12.1 nm)"]],ranks_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```

## M2

Using estimates from M2 age 60-69

```{r}


ranks_lt59_m2 <- all_strat_M2_msea %>% 
   filter(age_cat == "60-69")  %>%
   pull(estimate)

names(ranks_lt59_m2) <- all_strat_M2_msea %>% filter(age_cat == "60-69") %>% pull(metabolite)

fgsea_res_m2_lt69 <- fgsea(
  pathways = test,
  stats = ranks_lt59_m2
) %>%
  mutate(age_cat = "60-69",
         model = "M2")

head(fgsea_res_p1_lt50)

plotEnrichment(test[["Large HDL (average diameter 12.1 nm)"]],ranks_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```


# MSEA Age Category 70-85

## M1 

Using estimates from M1 age 70-85

```{r}


ranks_lt85_m1 <- all_strat_M1_msea %>% 
   filter(age_cat == "70-85")  %>%
   pull(estimate)

names(ranks_lt85_m1) <- all_strat_M1_msea %>% filter(age_cat == "70-85") %>% pull(metabolite)

fgsea_res_m1_lt85 <- fgsea(
  pathways = test,
  stats = ranks_lt85_m1
) %>%
  mutate(age_cat = "70-85",
         model = "M1")

head(fgsea_res_p1_lt50)

plotEnrichment(test[["Large HDL (average diameter 12.1 nm)"]],ranks_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```

## M2

Using estimates from M2 age 70-85

```{r}


ranks_lt85_m2 <- all_strat_M2_msea %>% 
   filter(age_cat == "70-85")  %>%
   pull(estimate)

names(ranks_lt85_m2) <- all_strat_M2_msea %>% filter(age_cat == "70-85") %>% pull(metabolite)

fgsea_res_m2_lt85 <- fgsea(
  pathways = test,
  stats = ranks_lt85_m2
) %>%
  mutate(age_cat = "70-85",
         model = "M2")

head(fgsea_res_p1_lt50)

plotEnrichment(test[["Large HDL (average diameter 12.1 nm)"]],ranks_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```


```{r}
all_msea_res <- bind_rows(list(fgsea_res_m1_lt50, fgsea_res_m1_lt59, fgsea_res_m1_lt69, fgsea_res_m1_lt85, fgsea_res_m2_lt50, fgsea_res_m2_lt59, fgsea_res_m2_lt69, fgsea_res_m2_lt85))
```

```{r}
# forest for msea 

# trying loop to get plot for all the subgroups

met_cats <- unique(all_msea_res$pathway)

models <- unique(all_msea_res$model)

map_res <- data.frame(
  models = models) %>%
  mutate(data = map(models, ~ all_msea_res %>% 
                      filter(model == .x)%>%
                      arrange(NES) %>%
                      complete(age_cat,pathway,
                               fill = list(NES = 0))))
library(forcats)

map2(.x = map_res$data, .y = models, 
                        ~ggplot(.x,
       aes(x = pathway, y = NES, col = age_cat, fill = age_cat)) +
        geom_col(position = "dodge") +
        coord_flip() +
        theme_minimal() +
        theme() +
        ggtitle(.y))

test <- all_msea_res %>%
 # arrange(NES) %>%
  #filter(padj <0.05)%>%
                      complete(age_cat,pathway,model,
                               fill = list(NES = 0)) %>%
  mutate(pathways = fct_relevel(pathway, topPathways)) %>%
  mutate(significant = ifelse(padj <0.05, 1, 0))

ggplot(test,
       aes(x = pathways, y = NES, col = age_cat, fill = age_cat)) +
        geom_col(position = "dodge") +
        geom_text(aes(label = ifelse(significant, "*", ""), group = age_cat), 
            position = position_dodge(width = .9), size = 20 / .pt)+
        coord_flip() +
        theme_minimal() +
        theme() +
        facet_grid(~model)
```

