---
title: "MSEA"
author: "Margaret Janiczek"
date: "2024-01-03"
output:
  html_document:
    df_print: paged
    toc: TRUE
    toc_float: TRUE
    number_sections: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
#BiocManager::install("fgsea")
library(fgsea)
library(ggplot2)
library(tidyr)
library(dplyr)
```

```{r}
# example
data(examplePathways)
data(exampleRanks)
set.seed(42)

fgseaRes <- fgsea(pathways = examplePathways,
                  stats = exampleRanks,
                  minSize = 15,
                  maxSize = 500)
```

```{r}
#now load metabolite group info

#
load("processed_met_key.Rdata")

library(stringr)

met_key <- ukb_key_metab %>%
  filter(!str_detect(col.name, "qc_flag"))%>%
  filter(field.showcase %in% 23400:24000) %>%
  filter(col.type == "Continuous") %>%
  filter(str_detect(field.tab, ".0.0")) %>%
  mutate(pathway = ifelse(
    Group == "Lipoprotein subclasses", Subgroup,Group
  ))

met_key$pathway[is.na(met_key$pathway)] <- "Other"

# make list of mets per met group
test <- split(met_key$col.name,met_key$pathway)

res81 <- read.csv("validated_unique_mets_81_details.csv")

res812 <- read.csv("results_mod1_bothphases.csv") %>%
  #filter(metabolite %in% res81$metabolite) %>%
  arrange(estimate)

load("results_mod2_bothphases.Rdata")
```

# MSEA Age Category <50

## M1 P1

Using estimates from Phase 1 M1 age <50

```{r}


ranks_lt50 <- res812 %>% 
   filter(age_cat == "<50" & phase == "Phase I")  %>%
   pull(estimate)

names(ranks_lt50) <- res812 %>% filter(age_cat == "<50" & phase == "Phase I") %>% pull(metabolite)

fgsea_res_p1_lt50 <- fgsea(
  pathways = test,
  stats = ranks_lt50
)

head(fgsea_res_p1_lt50)

plotEnrichment(test[["Large HDL (average diameter 12.1 nm)"]],ranks_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```

## M2 P1

Using estimates from Phase 1 M2 age <50

```{r}


ranks_lt50 <- all_strat_both_phases_long_m2 %>% 
   filter(age_cat == "<50" & phase == "Phase I")  %>%
   pull(estimate)

names(ranks_lt50) <- all_strat_both_phases_long_m2 %>% filter(age_cat == "<50" & phase == "Phase I") %>% pull(metabolite)

fgsea_res_p1_lt50 <- fgsea(
  pathways = test,
  stats = ranks_lt50
)

head(fgsea_res_p1_lt50)

plotEnrichment(test[["Large HDL (average diameter 12.1 nm)"]],ranks_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```

## M1 P1

Using estimates from Phase 1 M1 age <50

```{r}


ranks_lt50 <- res812 %>% 
   filter(age_cat == "<50" & phase == "Phase II")  %>%
   pull(estimate)

names(ranks_lt50) <- res812 %>% filter(age_cat == "<50" & phase == "Phase II") %>% pull(metabolite)

fgsea_res_p1_lt50 <- fgsea(
  pathways = test,
  stats = ranks_lt50
)

head(fgsea_res_p1_lt50)

plotEnrichment(test[["Large HDL (average diameter 12.1 nm)"]],ranks_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```

## M2 P2

Using estimates from Phase 2 M2 age <50

```{r}


ranks_lt50 <- all_strat_both_phases_long_m2 %>% 
   filter(age_cat == "<50" & phase == "Phase II")  %>%
   pull(estimate)

names(ranks_lt50) <- all_strat_both_phases_long_m2 %>% filter(age_cat == "<50" & phase == "Phase II") %>% pull(metabolite)

fgsea_res_p1_lt50 <- fgsea(
  pathways = test,
  stats = ranks_lt50
)

head(fgsea_res_p1_lt50)

plotEnrichment(test[["Large HDL (average diameter 12.1 nm)"]],ranks_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```


# MSEA Age Category 50-59

## M1 P1

Using estimates from Phase 1 M1 age 50-59

```{r}


ranks_lt50 <- res812 %>% 
   filter(age_cat == "50-59" & phase == "Phase I")  %>%
   pull(estimate)

names(ranks_lt50) <- res812 %>% filter(age_cat == "50-59" & phase == "Phase I") %>% pull(metabolite)

fgsea_res_p1_lt50 <- fgsea(
  pathways = test,
  stats = ranks_lt50
)

head(fgsea_res_p1_lt50)

plotEnrichment(test[["Large HDL (average diameter 12.1 nm)"]],ranks_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```

## M2 P1

Using estimates from Phase 1 M2 age 50-59

```{r}


ranks_lt50 <- all_strat_both_phases_long_m2 %>% 
   filter(age_cat == "50-59" & phase == "Phase I")  %>%
   pull(estimate)

names(ranks_lt50) <- all_strat_both_phases_long_m2 %>% filter(age_cat == "50-59" & phase == "Phase I") %>% pull(metabolite)

fgsea_res_p1_lt50 <- fgsea(
  pathways = test,
  stats = ranks_lt50
)

head(fgsea_res_p1_lt50)

plotEnrichment(test[["Large HDL (average diameter 12.1 nm)"]],ranks_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```

## M1 P1

Using estimates from Phase 1 M1 age 50-59

```{r}


ranks_lt50 <- res812 %>% 
   filter(age_cat == "50-59" & phase == "Phase II")  %>%
   pull(estimate)

names(ranks_lt50) <- res812 %>% filter(age_cat == "50-59" & phase == "Phase II") %>% pull(metabolite)

fgsea_res_p1_lt50 <- fgsea(
  pathways = test,
  stats = ranks_lt50
)

head(fgsea_res_p1_lt50)

plotEnrichment(test[["Large HDL (average diameter 12.1 nm)"]],ranks_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```

## M2 P2

Using estimates from Phase 2 M2 age 50-59

```{r}


ranks_lt50 <- all_strat_both_phases_long_m2 %>% 
   filter(age_cat == "50-59" & phase == "Phase II")  %>%
   pull(estimate)

names(ranks_lt50) <- all_strat_both_phases_long_m2 %>% filter(age_cat == "50-59" & phase == "Phase II") %>% pull(metabolite)

fgsea_res_p1_lt50 <- fgsea(
  pathways = test,
  stats = ranks_lt50
)

head(fgsea_res_p1_lt50)

plotEnrichment(test[["Large HDL (average diameter 12.1 nm)"]],ranks_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```

# MSEA Age Category 60-69

## M1 P1

Using estimates from Phase 1 M1 age 60-69

```{r}


ranks_lt50 <- res812 %>% 
   filter(age_cat == "60-69" & phase == "Phase I")  %>%
   pull(estimate)

names(ranks_lt50) <- res812 %>% filter(age_cat == "60-69" & phase == "Phase I") %>% pull(metabolite)

fgsea_res_p1_lt50 <- fgsea(
  pathways = test,
  stats = ranks_lt50
)

head(fgsea_res_p1_lt50)

plotEnrichment(test[["Large HDL (average diameter 12.1 nm)"]],ranks_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```

## M2 P1

Using estimates from Phase 1 M2 age 60-69

```{r}


ranks_lt50 <- all_strat_both_phases_long_m2 %>% 
   filter(age_cat == "60-69" & phase == "Phase I")  %>%
   pull(estimate)

names(ranks_lt50) <- all_strat_both_phases_long_m2 %>% filter(age_cat == "60-69" & phase == "Phase I") %>% pull(metabolite)

fgsea_res_p1_lt50 <- fgsea(
  pathways = test,
  stats = ranks_lt50
)

head(fgsea_res_p1_lt50)

plotEnrichment(test[["Large HDL (average diameter 12.1 nm)"]],ranks_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```

## M1 P1

Using estimates from Phase 1 M1 age 60-69

```{r}


ranks_lt50 <- res812 %>% 
   filter(age_cat == "60-69" & phase == "Phase II")  %>%
   pull(estimate)

names(ranks_lt50) <- res812 %>% filter(age_cat == "60-69" & phase == "Phase II") %>% pull(metabolite)

fgsea_res_p1_lt50 <- fgsea(
  pathways = test,
  stats = ranks_lt50
)

head(fgsea_res_p1_lt50)

plotEnrichment(test[["Large HDL (average diameter 12.1 nm)"]],ranks_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```

## M2 P2

Using estimates from Phase 2 M2 age 60-69

```{r}


ranks_lt50 <- all_strat_both_phases_long_m2 %>% 
   filter(age_cat == "60-69" & phase == "Phase II")  %>%
   pull(estimate)

names(ranks_lt50) <- all_strat_both_phases_long_m2 %>% filter(age_cat == "60-69" & phase == "Phase II") %>% pull(metabolite)

fgsea_res_p1_lt50 <- fgsea(
  pathways = test,
  stats = ranks_lt50
)

head(fgsea_res_p1_lt50)

plotEnrichment(test[["Large HDL (average diameter 12.1 nm)"]],ranks_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```

# MSEA Age Category 70-85
## M1 P1

Using estimates from Phase 1 M1 age 70-85

```{r}


ranks_lt50 <- res812 %>% 
   filter(age_cat == "70-85" & phase == "Phase I")  %>%
   pull(estimate)

names(ranks_lt50) <- res812 %>% filter(age_cat == "70-85" & phase == "Phase I") %>% pull(metabolite)

fgsea_res_p1_lt50 <- fgsea(
  pathways = test,
  stats = ranks_lt50
)

head(fgsea_res_p1_lt50)

plotEnrichment(test[["Large HDL (average diameter 12.1 nm)"]],ranks_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```

## M2 P1

Using estimates from Phase 1 M2 age 70-85

```{r}


ranks_lt50 <- all_strat_both_phases_long_m2 %>% 
   filter(age_cat == "70-85" & phase == "Phase I")  %>%
   pull(estimate)

names(ranks_lt50) <- all_strat_both_phases_long_m2 %>% filter(age_cat == "70-85" & phase == "Phase I") %>% pull(metabolite)

fgsea_res_p1_lt50 <- fgsea(
  pathways = test,
  stats = ranks_lt50
)

head(fgsea_res_p1_lt50)

plotEnrichment(test[["Large HDL (average diameter 12.1 nm)"]],ranks_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```

## M1 P1

Using estimates from Phase 1 M1 age 70-85

```{r}


ranks_lt50 <- res812 %>% 
   filter(age_cat == "70-85" & phase == "Phase II")  %>%
   pull(estimate)

names(ranks_lt50) <- res812 %>% filter(age_cat == "70-85" & phase == "Phase II") %>% pull(metabolite)

fgsea_res_p1_lt50 <- fgsea(
  pathways = test,
  stats = ranks_lt50
)

head(fgsea_res_p1_lt50)

plotEnrichment(test[["Large HDL (average diameter 12.1 nm)"]],ranks_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```

## M2 P2

Using estimates from Phase 2 M2 age 70-85

```{r}


ranks_lt50 <- all_strat_both_phases_long_m2 %>% 
   filter(age_cat == "70-85" & phase == "Phase II")  %>%
   pull(estimate)

names(ranks_lt50) <- all_strat_both_phases_long_m2 %>% filter(age_cat == "70-85" & phase == "Phase II") %>% pull(metabolite)

fgsea_res_p1_lt50 <- fgsea(
  pathways = test,
  stats = ranks_lt50
)

head(fgsea_res_p1_lt50)

plotEnrichment(test[["Large HDL (average diameter 12.1 nm)"]],ranks_lt50)

topPathwaysUp <- fgsea_res_p1_lt50 %>%
  filter(ES >0) %>%
  arrange(pval) %>%
  pull(pathway)
  

topPathwaysDown <- fgsea_res_p1_lt50 %>%
  filter(ES <0) %>%
  arrange(pval) %>%
  pull(pathway)

topPathways <- c(topPathwaysUp, rev(topPathwaysDown))

plotGseaTable(test[topPathways], ranks_lt50, fgsea_res_p1_lt50, gseaParam = 0.5)
```


